
---
title: "Project3"
author: "Rongkui Han"
date: "2/6/2020"
output:
  bookdown::pdf_document2: default
  toc: no
---

```{r}
library(AER)
library(kableExtra)
library(knitr)
library(latexpdf)
library(ggplot2)
library(dplyr)

```

```{r}
data("Fatalities")
data = Fatalities

data = mutate(data, yearpast = (as.numeric(data$year))-1)
data = mutate(data, fatalratio = (fatal/pop)*100)

#Impute values for missing CA data
data[28,15] = as.factor("no")
data[28,16] = as.factor("no")
```


***

### Team ID: Team 6

#### NAME: Connor Rosenberg
#### NAME: Rongkui Han
#### NAME: Yuqing Yang
#### NAME: Nassim Ali-Chaouche

***

## 1.0 Introduction


#### 1.1 Background

#### 1.2 Questions of Interest

         

## 2.0 Analysis Plan


#### 2.1 Population and study design

#### 2.2 Statistical Analysis   

##### 2.2.1 Descriptive Analysis   



##### 2.2.2 Main Analysis


#### 2.3 Model Diagnostics


## 3.0 Results


#### 3.1 Descriptive Analysis     

#### 3.2 Main Analysis

#### 3.3 Model Diagnostics


#### 3.4 Causal Effects



## 4.0 Discussion


## 5.0 Reference



