---
title: "Project_3_Nassim"
author: "Nassim Ali-Chaouche"
date: "2/6/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(AER)
library(MatchIt)
library(dplyr)
library(ggplot2)
library(lme4)
library(car)
```

```{r}
data(Fatalities)

```

```{r}
?Fatalities
```

```{r}
View(Fatalities)
```


1.1 Background

The "Fatalities" data set contains data pertinent to US traffic fatalities from the years 1982 to 1988. The data set includes panel data for 48 states (data from Alaska and Hawaii are not included). The variables involved in the study include income per capita, a beer tax, a mandatory jail sentence for an initial drunk driving conviction, population, and the number of vehicle fatalites. The observations were recorded for each state annually. In total, there are 336 observations recorded for 34 distinct variables. The data was obtained from sources such as the US Department of Transportation Fatal Accident Reporting System and the US Bureau of Labor Statistics. 

Due to the observational nature of the data, obtaining causal effects may pose to be a challenge. In this report, we will attempt to discover if there is a causal relationship between having a mandatory jail sentence for an initial drunk driving conviction and the number of traffic fatalities. In particular, we will study if a mandatory jail sentence is related to having a lower number of traffic fatalites. 


```{r}
data("Fatalities")
data = Fatalities

#Impute values for missing CA data
data[28,15] = as.factor("no")
data[28,16] = as.factor("no")


#head(data)
#dim(data)
data['fr'] = data$fatal/data$pop*10000
```

```{r}
mod_match = matchit(jail ~ year + spirits + unemp + income + beertax + baptist + mormon + drinkage + dry + youngdrivers + miles + breath + pop + gsp, method = "nearest", data = data)

dta_m = match.data(mod_match)
```



# Model Diagnostics

The two crucial assumptions for a mixed effect model are normally distributed residuals and constant variance of the residuals across the fitted values, which will be discussed below. 


```{r, include = FALSE}
# Mixed effects model
fit2 <- lmer(fr ~ jail + (1 | state), data = dta_m)
summary(fit2)
```



```{r, echo = FALSE, fig.align="center"}
qqnorm(residuals(fit2))
qqline(residuals(fit2), col = "red")
```


```{r, echo = FALSE, fig.align="center"}
hist(residuals(fit2), main = "Histogram of Residuals", xlab = "Residual Values")
```



```{r, echo = FALSE, fig.align="center"}
plot(fit2)
```



```{r, echo = FALSE, fig.align="center"}
hist(fit2$residuals, main = "Histogram of Residuals", xlab = "Residual Values")
```

Figure ..: Visual diagnostics of Mixed Effect model assumptions. (a). Normal Q-Q plot of residuals. (b) Histogram of model residuals. (c) Residuals-versus-fitted value scatter plot. 




#####  Normality:

From the Q-Q plot of the residuals (Figure ..a), we can see that the probability mass on the left and right tails are higher than what is expected from a normal distribution. The distribution of the residuals seem to be heavy-tailed. Thus, the normality assumption is not satisfied from the Q-Q plot. A histogram is used to visualize the distribution of the residuals (Figure ..b). From the histogram, it is not easily noticeable that the residuals do not follow a normal distribution. 

To further test for normality of the errors, a Shapiro-Wilk test will be used on the distribution of the residuals. A Shapiro-Wilk test is used to test whether a distribution of data follows a normal distribution.

The null and alternative hypotheses of the Shapiro-Wilk test are:   
$H_0$: The residuals are normally distributed.   
$H_1$: The residuals are not normally distributed.   

```{r, include = FALSE}
shapiro.test(residuals(fit2))
```

The p-value is essentially 0, and thus we reject the null hypothesis. Thus, there is evidence that the distribution of the residuals do not follow a normal distribution.


##### Equal Variances:

The spread of the residuals seem to increase along the x-axis in the residuals-versus-fitted value scatter plot (Figure ..c), indicating unequal variance of the residuals across the fitted values. Thus, the equal variance assumption is not satisfied. 



##### Response Variable Transformation:

In an attempt to remedy the departures of the assumptions of the mixed effect model, we used a log transformation of the response variable, fr. After the transformation, we concluded that the distribution of the residuals did not follow a normal distribution after looking at a Normal Q-Q plot and performing a Shapiro-Wilk test. The residuals did seem to have more equal variance across the fitted values after the transformation. 

```{r, include = FALSE}
fit3 <- lmer(log(fr) ~ jail + (1 | state), data = dta_m)
```

```{r, include = FALSE}
qqnorm(residuals(fit3))
qqline(residuals(fit3), col = "red")
```

```{r, include = FALSE}
hist(residuals(fit3), main = "Histogram of Residuals", xlab = "Residual Values")
```


```{r, include = FALSE}
plot(fit3)
```

```{r, include = FALSE}
shapiro.test(residuals(fit3))
```








