---
title: "Project3"
author: "Rongkui Han"
date: "2/6/2020"
output:
  bookdown::pdf_document2: default
  toc: no
---

```{r}
library(AER)
library(sqldf)
library(kableExtra)
library(knitr)
library(latexpdf)
library(ggplot2)

```


Make suggestions to policymakers to take certain measures by discovering variables that caused the reduction or increase of traffic fatalities.    

```{r}

data("Fatalities")
data = Fatalities

```


*Question:* Which factor variables change across time within each state?

```{r}
factorchange_table = data.frame(State = c("CO","CT","IL","IN","IA","KS","KY","LA","MS","NV","NH","OH","OR","UT"),
                                Breath = c("1982-1983","NA","1986-1987","1983-1984","1982-1983","1985-1986","1983-1984","1982-1983","1982-1983","NA","1982-1983","NA","NA","NA"),
                                Jail = c("NA","1984-1985","NA","NA","NA","NA","NA","NA","NA","1982-1983","NA","1982-1983 & 1986-1987","1983-1984","1982-1983"),
                                Service = c("NA","1984-1985","NA","NA","NA","NA","NA","NA","NA","1982-1983","NA","NA","1983-1984","1982-1983")
                                )

kable(factorchange_table)
```


```{r}

factorpercent_table = data.frame(Year = c("1982","1983","1984","1985","1986","1987","1988"),
                                 Breath = c("37.5%","45.8%","45.8%","45.8%","47.9%","50.0%","50.0%"),
                                 Jail = c("18.8%","27.0%","29.2%","31.3","31.3%","29.2%","29.2%"),
                                 Service = c("10.4%","16.7%","18.8%","20.8%","20.8%","20.8%","20.8%"))
kable(factorpercent_table)

```



```{r}

data = data[complete.cases(data),]
dim(data)
ggplot(data, aes(x = jail, y = fatal)) +
  geom_boxplot()


```

```{r}
data['fr'] = data$fatal/data$pop*10000

ggplot(data, aes(x = jail, y = fr)) +
  geom_boxplot()

```


```{r}
ggplot(data, aes(x = drinkage, y = fr)) +
  geom_point()
```


1. Explore this dataset and generate summary statistics (in forms of tables or plots) that you find crucial for your own interest, or for convincing the policymakers.

2. Consider only the full dataset from 1982 to 1988, propose a regression model to study whether having a mandatory jail sentence is associated with reduced traffic fatalities. In particular, you need to
- specify your model,

*Longitudinal Analysis Model*
$$Y_{i,T} = \beta_1X_{1,i,T} + ... +\beta_kX_{k,i,T} + \alpha_i + \lambda_T + \epsilon_{i,T}$$
$\alpha$ is the fixed effect due to each state - binary
$\lambda$ is the fixed effect due to each time period - biary or continuous
$\beta_kX_{k,i,T}$ is the effect of the various dependent variables

- state the assumptions required,

1) Residuals are IID, have mean zero & constant variance.
2) No large outliters
3) No multicolinearity


- fit the model with appropriate methods,
- conduct model diagnostics and/or sensitivity analysis,
- and discuss causal interpretation of the proposed models.

3. Conclude your analysis results. You may want to test a hypothesis, construct a confidence interval, or draw a confidence band.

4. Explain the implications of your results to policymakers who know little about statistics. Make suggestions if you want to.



