---
title: "Project3"
output:
  bookdown::pdf_document2:
    toc: no
---

<style type="text/css">

body{ /* Normal  */
      font-size: 18px;
  }

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE,message=FALSE,warning=FALSE)
```

# Diagnostics for Propensity Score matching

After the propensity score matching, we would check whether our propensity score model has been adequately specified using the following steps:

(1) we would assess the standardized differences of each covariate between treatment groups. 
(2) we would assess the distributions of propensity scores by treatment.
(3) we would compare distributions of the covariates between treatments.

## Result

(1) The standardized differences of each covariate between treatment groups.

Table \@ref(tab:stdf) shows the results of standardized differences and the 95% confidence interval for covariates. It can be seen that all the confidence intervals include 0, we are 95% confidence that there is no differences between treatment groups for covariates.  

```{r}
library("AER")
library(tidyverse)
data(Fatalities)

library("directlabels")
library(gridExtra)

data = Fatalities %>% 
  mutate(fatal_rate = fatal/pop*10000)

state_1 <- c('ca', 'ct', 'nv', 'oh', 'or', 'sc', 'ut')

#data[data$state == "ca" & data$year == "1988",]$jail = "no"
data <- data %>% drop_na()

library(MatchIt)
data$jail = ifelse(data$jail == 'yes', 1, 0)
#data$service = ifelse(data$service == 'yes', 1, 0)
#data$breath = ifelse(data$breath == 'yes', 1, 0)

data_cov = c("spirits" , "unemp" , "income" , "beertax" , "baptist" , "mormon" , "drinkage" , "dry" , "youngdrivers" , "miles" , "pop" , "gsp")

mod_match = matchit(jail ~  spirits + unemp + income + beertax + baptist + mormon + drinkage + dry + youngdrivers + miles + breath + pop + gsp, method = "nearest", data = data)

dta_m = match.data(mod_match)
```

```{r stdf}
library(stddiff)
library(cobalt)

#bal.tab(mod_match, m.threshold = .1)
stdf <- stddiff.numeric(dta_m, 15, c(3:13, 27, 34))
colnames(stdf) <- c("Mean(jail = 0)","Sd(jail = 0)","Mean(jail = 1)","Sd(jail = 1)","missing.c","missing.t", "Standardized difference",  "Lower bound", "Upper bound")
stdf <- stdf %>% as.tibble()
knitr::kable(stdf[,c(1:4,7:9)], caption = "Standardized differences of covariates between treatment groups.")
```

(2) Distributions of propensity scores by treatment.
(3) Distributions of the covariates between treatments.

Figure \@ref(fig:cov_bp) shows that distributions of the covariates are similar between treatments.
```{r cov_bp, fig.cap="Distributions of the covariates "}
par(mfrow = c(3,4))
for (i in 1:length(data_cov)){
  boxplot(dta_m[[data_cov[i]]]~dta_m$jail, xlab = "Jail", ylab = data_cov[i])
}
par(mfrow = c(1,1))
```

